---
title: "Groups of Data - Vectors, Matrices, Lists & Dataframes"
author: Mostafa S. Abd El-Maksoud
output: html_notebook
---

In this notebook, we will learn how to handle different types of data structure.

R provides several ways for grouping objects: - **Vector**: Collection of objects with the same data type. - **Matrix**: Table consisting of objects with the same data type. - **List**: Collection of objects with possibly different data types. - **Dataframes**: A table or 2D structure where each column can be of different type (like a spreadsheet).

## Vector

-   A vector is a collection of objects with the same data type.
-   To create a vector, use the `c()` function:

```{r}
amino_acids <- c("methionine", "leucine", "alanine", "valine", "glutamine", "threonine")
```

-   Accessing items in a vector can be done like so:

```{r}
amino_acids[1] # Access the first element
amino_acids[2] # Access the second element
```

-   We can also combine two vectors into a single vector using the `c()` function:

```{r}
add_amino_acids <- c("glycine", "cysteine", "serine")
many_amino_acids <- c(amino_acids, add_amino_acids)
many_amino_acids
```

## Matrix

-   A matrix is a table consisting of objects with the same data type.
-   To create a matrix, use the `matrix()` function:

```{r}
amino_acids_matrix <- matrix(c("methionine", "leucine", "alanine", "valine", "glutamine", "threonine"), nrow=3, ncol=2)
amino_acids_matrix
```

-   We can add rows using `rbind()`:

```{r}
amino_acids_matrix <-  rbind(amino_acids_matrix, c("proline", "arginine"))
amino_acids_matrix
```

-   Also we can add columns using `cbind()`:

```{r}
amino_acids_matrix <-  cbind(amino_acids_matrix, c("histidine", "phenylalanine", "tryptophan", "selenocysteine"))
amino_acids_matrix
```

-   Accessing items in a matrix can be done like so:

```{r}
amino_acids_matrix[1, 2] # Accesses the first row, second column
amino_acids_matrix[4, 3] # Accesses the fourth row, third column
```

-   We can also check if the items in a matrix satisfies a given condition:
-   `any()` checks if at least one of the items satisfies the condition
-   `all()` checks if all of the items satisfy the condition

```{r}
numbers <- matrix(c(2, 4, 6, 8, 10, 12, 14, 16, 18), nrow=3, ncol=3)
any(numbers < 6)
any(numbers < 1)
all(numbers < 20)
all(numbers < 5)
```

-   It is possible to name the rows and columns (also called the *dimensions*) of a matrix using `colnames()` and `rownames()`, respectively.

-   This can greatly aid in making our matrix more descriptive, especially when we perform some analysis.

```{r}
colnames(numbers) <- c("Treatment 1", "Treatment 2", "Treatment 3")
rownames(numbers) <- c("Patient 1", "Patient 2", "Patient 3")
numbers
```

-   Finally, we demonstrate some matrix operations:

```{r}
numbers1 <- matrix(c(2, 4, 6, 8, 10, 12, 14, 16, 18), nrow=3, ncol=3)
numbers2 <- matrix(c(12, 14, 16, 18, 20, 22, 24, 26, 28), nrow=3, ncol=3)

numbers1 + numbers2
numbers1 - numbers2
numbers1 * numbers2
numbers1 / numbers2

100 * numbers1 # Scalar multiplication
t(numbers1) # Matrix transpose
```

## List

-   A list is a collection of objects with possibly different data types.
-   To create a list, use the `list()` function:

```{r}
assorted_list <- list("proline", "methionine", 1, 2)
assorted_list
```

-   Accessing list items can be a bit tricky though.
-   Observe how `assorted_list[1]` does not return the item `proline`. It actually returns a list containing `proline`.

```{r}
assorted_list[1]
```

-   If we want the item `proline` to be returned, we have to use double brackets:

```{r}
assorted_list[[1]]
```

-   If we want a summary of the list elements, we can use the `summary()` function.

```{r}
summary(assorted_list)
```

## Dataframe

-   In data analysis, information is often structured in organized tables that have a certain number of rows and columns like an Excel spreadsheet.

-   R data frames are a type of data structure designed to hold such tabular data. A data frame consists of a number of rows and columns, with each column representing some variable or feature of the data and each row representing a record.

-   A data frame is similar to a matrix in that it is a 2-dimensional data structure but unlike a matrix, different columns can hold data of different types. It is composed of many lists.

-   You can create a new data frame by passing vectors of the same length to the `data.frame()` function.

```{r}
a <- c(1, 2, 3, 4)
b <- c("R", "Python", "PHP", "JS")
c <- c(TRUE, FALSE, TRUE, FALSE)

# Create a new data frame
my_frame <- data.frame(a, b, c)

my_frame
```

-   You can also change your column and row names.

```{r}
# Change column names
colnames(my_frame) <- c("c1", "c2", "c3")

# Add row names
rownames(my_frame) <- c("r1", "r2", "r3", "r4")
my_frame
```

-   For reference, we will be loading a dataset on phage-host interaction to illustrate the use of `dataframes` in reading dataset files.

```{r}
dataset <- read.delim("data.tsv")
```

**Summarizing Data Frames**

-   Data frames support many of the summary functions that apply to matrices and lists.

-   The `str()` function gives an overview of the structure of the Data Frame. It is useful to check the structure first, since running a full summary can take a while if you are working with a lot of data.

-   The `summary()` function gives summary statistics for each variable in the data frame.

```{r}
str(dataset) # get a sense of the range of values in your dataset
summary(dataset)
```

-   Data frames support a few other basic summary operations.

```{r}
dim(dataset)
nrow(dataset)
ncol(dataset)
```

-   If a data frame is large, you won't want to try to print the entire frame to the screen. You can look at a few rows at the beginning or end of a data frame using the `head()` and `tail()` functions respectively:

```{r}
head(dataset, 5)

tail(dataset, 5)
```

**Dataframe Indexing**

-   Since data frame are lists under the hood, where each list object is a column, they support the indexing operations that apply to lists.

-   Data frames also support matrix-like indexing by using a single square bracket with a comma separating the index value for the row and column. Matrix indexing allows you get values by row or specific values within the data frame.

***You can also delete columns in a data frame by assigning them a value of NULL.***

```{r}
head(dataset$model)
head(dataset[["model"]])
head(dataset[[1]])

# Get the value at row 2 column 6
dataset[2, 6]

# Get the second row
dataset[2, ]

# Get the 6th column
dataset[, 6]

# Get a column by using its name
dataset[, "mpg"]
```

**Filtering of Dataframes**

-   You can also use the `subset()` function to create data frame subsets based on logical statements. `subset()` takes the data frame as the first argument and then a logical statement as the second argument create a subset.

```{r}
subset(dataset, (dataset$hp > 160))
```
