---
title: "Introduction to R - session 4"
Author: "Mostafa S. Abd El-Maksoud"
output: html_notebook

---

# Read Data

```{r}
# Read the data from a text file with headers
data = read.table("E:/courses/Introduction to R/Session 4/surgery.txt", header = T)
```


# Data inspection

```{r}
# Open a data viewer window to inspect the dataset
View(data)

# Display the structure of the dataset:
str(data)
```

---------------------------------------------------------------

# Data Preprocessing: Cleaning and Formatting the Surgery Dataset

```{r}
# Convert 'sex' from numeric codes to factor with labels
data$sex <- factor(data$sex, labels = c("female", "male"))

# Convert 'SES' (socioeconomic status) from numeric to labeled factor
data$SES <- factor(data$SES, labels = c("low", "middle", "high"))

# Convert 'complic' (complications) to yes/no factor
data$complic <- factor(data$complic, labels = c("no", "yes"))

# Convert 'comorb' (comorbidities) to yes/no factor
data$comorb <- factor(data$comorb, labels = c("no", "yes"))

# Convert 'depressi' (depression) to yes/no factor
data$depressi <- factor(data$depressi, labels = c("no", "yes"))

# Convert 'diabetes' to yes/no factor
data$diabetes <- factor(data$diabetes, labels = c("no", "yes"))

# Convert 'surgery' type from numeric code to descriptive factor
data$surgery <- factor(data$surgery, labels = c("gastric band", "gastric bypass"))

# Convert BMIpost, BMIpre, weight, and length to numeric
# Note : You should replace commas with dots before conversion
data$BMIpost <- as.numeric(gsub(",", ".", data$BMIpost))
data$BMIpre  <- as.numeric(gsub(",", ".", data$BMIpre))
data$weight  <- as.numeric(gsub(",", ".", data$weight))
data$length  <- as.numeric(gsub(",", ".", data$length))

# Check the updated structure of the data
str(data)
```

---------------------------------------------------------------

# Descriptive statistics

```{r}
# Show summary statistics for all variables in the dataset
summary(data)

# Calculate the mean (average) of 'weight', ignoring missing values
mean(data$weight, na.rm = TRUE)

# Calculate the median (middle value) of 'weight', ignoring missing values
median(data$weight, na.rm = TRUE)

# Get the minimum value of 'weight', ignoring missing values
min(data$weight, na.rm = TRUE)

# Get the maximum value of 'weight', ignoring missing values
max(data$weight, na.rm = TRUE)

# Get the 1st (25%) and 3rd (75%) quartiles of 'weight', ignoring missing values
quantile(data$weight, na.rm = TRUE, c(0.25, 0.75))
```


# Why do you need to add the argument na.rm=TRUE to these commands? What happens if you don’t?

- The argument `na.rm = TRUE` is used in functions like `mean()`, `median()`, `sum()`, to tell R to ignore any `NA` (missing) values during the calculation.

```{r}
# Example 1
median(c(10, 20, NA), na.rm = TRUE)
```

- in case of `na.rm = FALSE` R will give NA because it can't compute the function with a missing value unless you explicitly tell it to ignore NAs.

```{r}
# Example 2
median(c(10, 20, NA), na.rm = FALSE)
```


---------------------------------------------------------------

# For a categorical variable a frequency table is more appropriate

```{r}
# Frequency table for the 'sex' variable, Shows the count of males and females
table(data$sex)
```

- Calculate the mean age for females and males

```{r}
# Calculate the mean age for each sex group (female and male)
tapply(data$age, list(sex = data$sex), mean, na.rm = TRUE)
```

- Calculate correlation coefficient between weights and lengths

```{r}
# Load the 'corrplot' package for visualizing correlation matrices
library(corrplot)

# use = "complete.obs" means it ignores rows with missing values
cor(data$weight, data$length, use = "complete.obs")
```


---------------------------------------------------------------

# Several types of correlation coefficients exist. Go to the help of the cor() function and find which correlation coefficient is calculated by defaul

```{r}
?cor # or
help(cor)
# Answer -> (pearson)
```


# Add the appropriate argument to the cor() function in the previous command so that the Spearman correlation is calculated instead.

```{r}
cor(data$weight, data$length, use = "complete.obs", method = "spearman")
```

# Some more summary statistics

1. What is the maximum value of weight in the entire dataset?

```{r}
# Find the maximum value in the 'weight' column, ignoring missing values (NA)
max(data$weight, na.rm = TRUE)
```

2. How many missing values are there for age?

```{r}
# Count the number of missing values (NA) in the 'age' column
sum(is.na(data$age))
```

3. What is the mean value for BMIpre for the two surgical techniques (variable surgery)?

```{r}
# Calculate the mean of 'BMIpre' grouped by 'surgery' types
aggregate(BMIpre ~ surgery, data, mean, na.rm = TRUE)
```

4. What is the Pearson correlation between the BMI before the operation (BMIpre) and the BMI after the operation (BMIpost)?

```{r}
cor(data$BMIpre, data$BMIpost, use = "complete.obs", method = "pearson")
```


# 3. Graphics

1. Generate a bar chart of a categorical variable for the gender.

```{r}
barplot(table(data$sex), xlab = "Gender", ylab = "Frequency", main = "Gender Barplot")
```

2. Generate a bar chart graph with mean age in males and females.

```{r}
mean.age <- tapply(data$age, data$sex, mean, na.rm = TRUE)
barplot(mean.age, col = "skyblue", xlab = "Gender", ylab = "Mean Age", main = "Mean Age by Sex")
```

3. Make a histogram of a continuous variable “age”.

```{r}
hist(data$age, xlab = "Age", ylab = "Frequency", main = "Age Histogram")
```

4. Make a scatterplot of 2 continuous variables lengths and weights (using formula notation Y~X) with different colors for each SES-value

```{r}
plot(weight ~ length, data = data, col = data$SES, pch = 19, 
     main = "Weight vs Length by SES", xlab = "Length", ylab = "Weight")
legend("topright", legend = levels(colors), col = 1:length(levels(colors)), pch = 19)
```

5. Add the three regression lines for each of the 3 SES groups

```{r}
plot(weight ~ length, data = data, col = data$SES, pch = 19, 
     main = "Weight vs Length by SES", xlab = "Length", ylab = "Weight")

unique_ses <- unique(data$SES)
ses_colors <- rainbow(length(unique_ses))

# Add regression lines for each SES group
for(i in 1:length(unique_ses)) {
  ses_group <- unique_ses[i]
  subset_data <- data[data$SES == ses_group, ]
  abline(lm(weight ~ length, data = subset_data), col = ses_colors[i], lwd = 2)
}

# Add a legend to indicate the SES groups with corresponding colors
legend("topright", legend = levels(as.factor(data$SES)), col = ses_colors, 
       pch = 19, lwd = 2)

```

6. Make a boxplot of age and a separate boxplots per group we use the formula-notation (Y~X).

```{r}
# Set up the plotting area to display 2x2 grid of plots
par(mfrow = c(2,2))

# Create a boxplot of 'age', showing the distribution of age values
boxplot(data$age, 
        main = "Box plot of age",    # Title of the plot
        ylab = "Age")                # Label for the y-axis

# Create a boxplot of 'age' grouped by 'SES' (Socioeconomic Status)
boxplot(data$age ~ data$SES, 
        xlab = "SES group",          # Label for the x-axis (SES group)
        ylab = "Age",                # Label for the y-axis
        main = "Boxplot of Age by SES Group") # Title of the plot

# Create a boxplot of 'age' grouped by 'surgery' type (gastric band vs. gastric bypass)
boxplot(data$age ~ data$surgery, 
        xlab = "Surgery group",      # Label for the x-axis (Surgery group)
        ylab = "Age",                # Label for the y-axis
        main = "Boxplot of Age by Surgery Group")  # Title of the plot

# Create a boxplot of 'age' grouped by 'sex' (female vs. male)
boxplot(data$age ~ data$sex, 
        xlab = "Gender",             # Label for the x-axis (Gender)
        ylab = "Age",                # Label for the y-axis
        main = "Boxplot of Age by Gender")  # Title of the plot
```

7. Make a histogram of BMIpre.

```{r}
hist(data$BMIpre, xlab = "BMIpre", ylab = "Frequency", main = "Histogram of BMIpre")
```

8. Make a bar graph of the socio-economic status (SES).

```{r}
barplot(table(data$SES), xlab = "SES", ylab = "Frequency", main = "Barplot of socio-economic status", col = "skyblue")
```

9. Make a boxplot of BMIpre for the 2 surgical intervention types defined by surgery.

```{r}
boxplot(data$BMIpre ~ data$surgery, 
        xlab = "Surgical Intervention Type",  # Label for the x-axis
        ylab = "BMI Before Surgery",          # Label for the y-axis
        main = "Boxplot of BMI Before Surgery by Intervention Type",  # Plot title
        col = c("skyblue", "lightpink")       # Colors for each surgery type (skyblue for gastric band, lightpink for gastric bypass)
        # Uncomment the following line to set the y-axis limits and remove outliers
        # , ylim = c(1, 60)                    # Adjust the y-axis to exclude extreme outliers
)
```

10. Make a boxplot of BMIpost for the 2 surgical intervention types defined by surgery.

```{r}
boxplot(data$BMIpost ~ data$surgery, 
        xlab = "Surgical Intervention Type",  # Label for the x-axis
        ylab = "BMI Post Surgery",            # Label for the y-axis
        main = "Boxplot of BMI Post Surgery by Intervention Type",  # Plot title
        col = c("blue", "pink")               # Colors for each surgery type (blue for gastric band, pink for gastric bypass)
        # Uncomment the following line to set the y-axis limits and remove outliers
        # , ylim = c(1, 60)                    # Adjust the y-axis to exclude extreme outliers
)
```

11. Make a scatterplot of BMIpre (Y-axis) versus age.

```{r}
plot(x = data$age, y = data$BMIpre, xlab = "Age",
     ylab = "BMI Before Surgery", main = "Scatterplot of BMIpre vs Age",
     pch = 19, col = "darkgreen")
```

12. Make a scatterplot of BMIpre (Y-axis) versus age, using different markers for males and females.

```{r}
# Define point characters (pch) for each level of 'sex' (1 for female, 2 for male)
pch.values <- c(1, 2)
names(pch.values) <- levels(data$sex)

# Create a scatter plot of 'BMIpre' vs. 'age', with different colors and point shapes by 'sex'
# 'col' assigns a color based on 'sex' (converted to numeric), 'pch' assigns point shapes
plot(data$BMIpre ~ data$age, 
     col = as.numeric(data$sex),         # Color points based on 'sex' (1 for female, 2 for male)
     pch = pch.values[data$sex],         # Point shape based on 'sex' (1 for female, 2 for male)
     xlab = "Age",                       # X-axis label
     ylab = "BMI Before Surgery",        # Y-axis label
     main = "BMIpre vs Age by Sex")      # Title of the plot

# Add a legend to the top-right corner
legend("topright", 
       legend = levels(data$sex),        # Labels for 'sex' (female, male)
       col = 1:length(levels(data$sex)), # Color for each level of 'sex'
       pch = 1:length(levels(data$sex))) # Point shape for each level of 'sex'
```

- Add a line to the plot for each group

```{r}
# Plot
plot(data$BMIpre ~ data$age, col = as.numeric(data$sex), xlab = "Age", 
     ylab = "BMI Before Surgery", main = "BMIpre vs Age by Sex", 
     pch = as.numeric(data$sex))

# Add regression lines
for (gender in levels(data$sex)) {
  sub_data <- subset(data, sex == gender)
  model <- lm(BMIpre ~ age, data = sub_data)
  abline(model, col = which(levels(data$sex) == gender), lwd = 2)
}

# Define point shapes
pch.values <- c(1, 2)
names(pch_values) <- levels(data$sex)

# Add legend
legend("topright",
       legend = levels(data$sex),
       pch = pch.values,
       col = 1:length(levels(data$sex)),
       title = "Sex",
       pt.cex = 1.2)
```


# Outlier detection

- You should work on the dataset with this outlier corrected to 29.41.

```{r}
# Define the outlier threshold
outlires_threshold <- 40

# Replace values greater than the threshold in 'BMIpre' with 29.41
data$BMIpre[data$BMIpre > outlires_threshold] <- 29.41

# Replace values greater than the threshold in 'BMIpost' with 29.41
data$BMIpost[data$BMIpost > outlires_threshold] <- 29.41

# Save the modified dataset to a new file in tab-separated format
# row.names = FALSE excludes row names, col.names = TRUE includes column names in the output file
write.table(data, "E:/courses/Introduction to R/Session 4/modified_surgery.txt", sep = "\t", row.names = FALSE, col.names = TRUE)
```

# Testing for normality

- Check the normality using, histograms, QQ plots and Shapiro’s test, for BMIpre vs BMIpost
- Histogram for BMIpre and BMIpost

```{r}
par(mfrow = c(1,2))
hist(data$BMIpre, main = "Histogram of BMIpre", xlab = "BMIpre", col = "lightblue")
hist(data$BMIpost, main = "Histogram of BMIpost", xlab = "BMIpost", col = "lightgreen")
```

- QQ plots for BMIpre and BMIpost

```{r}
par(mfrow = c(1,2))
qqnorm(data$BMIpre, main = "QQ Plot of BMIpre")
qqline(data$BMIpre, col = "red")

qqnorm(data$BMIpost, main = "QQ Plot of BMIpost")
qqline(data$BMIpost, col = "blue")
```

- Shapiro test

```{r}
shapiro_pre <- shapiro.test(data$BMIpre)
shapiro_post <- shapiro.test(data$BMIpost)
print(shapiro_pre)
print(shapiro_post)
```