---
title: "Data visualization Part-1 (Base R Plotting)"
Author: "Mostafa S. Abd El-Maksoud"
output: html_notebook
---

# 1. bar plot

- A bar plot is a graphical display of data using bars of different heights. In R, the `barplot()` function is commonly used to visualize categorical data with corresponding numerical values. This type of chart is useful for comparing quantities across categories.

- In the example below, we will create a bar plot using a subset of the built-in `VADeaths` dataset, which contains death rates (per 1000 population) in Virginia, categorized by age group, gender, and area (rural or urban). We'll specifically visualize the death rates for Rural Males in the first three age groups.

```{r}
# Load the dataset and extract data for plotting
VADeaths                           # View the dataset
x <- VADeaths[1:3, "Rural Male"]   # Extract first 3 age groups for Rural Males
x                                  # View extracted values
```

- Basic Bar plots

```{r}
# Vertical bar plot (default)
barplot(x)

# Horizontal bar plot
barplot(x, horiz = TRUE)
```

- Change group names

```{r}
# Custom labels for each bar
barplot(x, names.arg = c("A", "B", "C"))
```

- Change color

```{r}
# Use a single color fill and a custom border color
barplot(x, names.arg = c("A", "B", "C"), col = "white", border = "steelblue")

# Use different border colors for each bar
barplot(x, col = "white", border = c("grey", "yellow", "blue"))

# Fill bars with a single color
barplot(x, col = "black")

# Fill bars with multiple colors
barplot(x, col = c("grey", "yellow", "blue"))

# Customize both labels and colors
barplot(x, names.arg = c("Group_A", "Group_B", "Group_C"),
        col = c("grey", "blue", "yellow"), border = "black")
```

- Change main title and axis labels

```{r}
# Add title and axis labels
barplot(x, main = "Death Rates in Virginia",
        xlab = "Age", ylab = "Rate")
```

- Stacked bar plots

```{r}
# Grouped (side-by-side) bars
# Stacked bar plot from the full VADeaths matrix
barplot(VADeaths,
        col = c("lightblue", "mistyrose", "lightcyan", "lavender", "cornsilk"),
        legend = rownames(VADeaths))
```

- Grouped bar plots

```{r}
# Grouped (side-by-side) bars
barplot(VADeaths,
        col = c("lightblue", "mistyrose", "lightcyan", "lavender", "cornsilk"),
        legend = rownames(VADeaths),
        beside = TRUE)
```

- Full Custom Grouped Bar Plot with Legend

```{r}
# Define a set of colors
# Define color palette
my_colors <- c("lightblue", "mistyrose", "lightcyan", "lavender", "cornsilk")

# Create grouped bar plot with custom options
barplot(VADeaths, col = my_colors, border = "black",
        beside = TRUE, horiz = FALSE,
        main = "Bar Plot", xlab = "X-axis", ylab = "Y-axis")

# Add legend at the top-left
legend("topleft", legend = rownames(VADeaths),
       fill = my_colors, box.lty = 1, cex = 0.8)
```


# 2. box plot

- Box plots (also called box-and-whisker plots) are useful for visualizing the distribution of numerical data across different groups. 
- They display the median, quartiles, and potential outliers of a dataset. 
- In R, the `boxplot()` function (and `gplots::boxplot2()`) can be used to easily create box plots with customizable styles and groupings.

```{r}
# Load the built-in dataset
data = ToothGrowth

# Display the first 6 rows of the dataset
head(data, 6)

# Show the structure of the dataset
str(data)
```

- R base box plots: boxplot()

```{r}
# Box plot of a single variable (tooth length)
boxplot(ToothGrowth$len)

# Box plots grouped by dose
boxplot(len ~ dose, data = ToothGrowth, frame = FALSE)

# Horizontal box plots
boxplot(len ~ dose, data = ToothGrowth, frame = FALSE,
        horizontal = TRUE)

# Notched box plots to compare medians
boxplot(len ~ dose, data = ToothGrowth, frame = FALSE,
        notch = TRUE)
```

- Change group names

```{r}
# Customize group labels
boxplot(len ~ dose, data = ToothGrowth, frame = FALSE,
        names = c("D0.5", "D1", "D2"))
```

- Change color

```{r}
# Change the border color of all boxes to one color
boxplot(len ~ dose, data = ToothGrowth, frame = FALSE,
        border = "steelblue")

# Use different border colors for each group
boxplot(len ~ dose, data = ToothGrowth, frame = FALSE,
        border = c("#999999", "#E69F00", "#56B4E9"))

# Fill all boxes with a single color
boxplot(len ~ dose, data = ToothGrowth, frame = FALSE,
        col = "steelblue")

# Use different fill colors for each group
boxplot(len ~ dose, data = ToothGrowth, frame = FALSE,
        col = c("#999999", "#E69F00", "#56B4E9"))
```

- Box plot with multiple groups

```{r}
# Create grouped box plots for interaction of two factors: supp and dose
boxplot(len ~ supp*dose, data = ToothGrowth,
        col = c("white", "steelblue"), frame = FALSE)
```

- Change main title and axis labels

```{r}
# Add main title, axis labels, and customize fill color
boxplot(len ~ dose, data = ToothGrowth,
        main = "Plot of Length by Dose",
        xlab = "Dose (mg)", ylab = "Tooth Length",
        col = "lightgray", frame = FALSE)
```

- Box plot with the number of observations: gplots::boxplot2()

```{r}
# Load gplots package for enhanced boxplot
library("gplots")

# Create a box plot with count annotations at the bottom
boxplot2(len ~ dose, data = ToothGrowth, frame = FALSE)

# Put count annotations at the top
boxplot2(len ~ dose, data = ToothGrowth, frame = FALSE, top = TRUE)
```

- plot multiple plots in one graph

```{r}
# Set layout to 2 rows and 2 columns
par(mfrow = c(2, 2))

# Plot 1: Step plot
plot(1:10, type = "S", main = "Plot 1")

# Plot 2: Histogram of normally distributed random numbers
hist(rnorm(100), main = "Plot 2")

# Plot 3: Box plot from mtcars dataset
boxplot(mpg ~ cyl, data = mtcars, main = "Plot 3")

# Plot 4: Bar plot of iris species count
barplot(table(iris$Species), main = "Plot 4")
```

# 3. Line Plots - R Base Graphs

- Line plots are used to display trends over time or ordered categories. 
- R’s base plot() and lines() functions allow flexible line plotting with custom point markers, line types, and colors.

```{r}
# Create some numeric data
x <- 1:10             # x-axis values
y1 <- x * x           # first y-values: square of x
y2 <- 2 * y1          # second y-values: twice y1

# Basic line plot using stair-step type
plot(x, y1, type = "S", main = "Stair Step Plot", xlab = "x", ylab = "y1")

# Show both points and connecting lines with custom color and point shape
plot(x, y1, type = "b", pch = 19, col = "red", xlab = "x", ylab = "y1")
```

- Plot Multiple Lines on the Same Graph

```{r}
# Plot the first line with points
plot(x, y1, type = "b", frame = TRUE, pch = 19, 
     col = "red", xlab = "x", ylab = "y-values",
     main = "Multiple Line Plot")

# Add the second line on the same graph
lines(x, y2, type = "b", pch = 18, col = "blue")
```

# 4. Histogram Plots in R

- A histogram shows the distribution of a numeric variable by dividing the data into bins (intervals) and counting how many observations fall into each bin.

- Create Sample Data

```{r}
# Set seed for reproducibility
set.seed(1234)

# Generate 400 weight values for two groups
x <- c(rnorm(200, mean = 55, sd = 5),    # Group 1 (e.g., Female)
       rnorm(200, mean = 65, sd = 5))    # Group 2 (e.g., Male)

# Preview first few values
head(x)
```

- Basic Histogram using hist()

```{r}
# Basic histogram with default settings
hist(x, col = "steelblue", 
     main = "Histogram of Weights",
     xlab = "Weight", ylab = "Frequency")
```

- Adjust the Number of Bins with breaks

```{r}
# Histogram with more bins (breaks = 30)
hist(x, col = "steelblue", 
     breaks = 30,
     main = "Histogram with 30 Bins",
     xlab = "Weight", ylab = "Frequency")
```

- You can also use other break rules such as "Sturges", "Scott", or "FD":

```{r}
# Histogram using the Sturges method (default)
hist(x, breaks = "Sturges", col = "skyblue",
     main = "Histogram (Sturges Method)",
     xlab = "Weight")
```

# 5. density plots: density()

- A density plot is a smoothed version of a histogram, estimating the probability distribution of a continuous variable using kernel density estimation.

- Basic Density Plot

```{r}
# Compute the kernel density estimate for miles per gallon (mpg)
dens <- density(mtcars$mpg)

# Basic density plot
plot(dens, 
     frame = FALSE, 
     col = "steelblue", 
     main = "Density Plot of mpg",
     xlab = "Miles Per Gallon (mpg)")
```

- Fill the Density Area using polygon()

```{r}
# Plot the density curve
plot(dens, 
     frame = FALSE, 
     col = "steelblue", 
     main = "Filled Density Plot of mpg",
     xlab = "Miles Per Gallon (mpg)")

# Fill the area under the curve
polygon(dens, col = "steelblue", border = "black")
```

