---
title: "Data visualization Part- (ggplot2)"
Author: "Mostafa S. Abd El-Maksoud"
output: html_notebook
---

- Make sure that you have both libraries installed. If not, you can install them first using:

```{r}
# Installing the libraries if not installed already
install.packages("ggplot2")
install.packages("ggsci")
```

- To start using the libraries you mentioned in your R code, you can import them with the library() function. Here's how you can load the libraries:

```{r}
# Importing necessary libraries
library(ggplot2)  # For data visualization using ggplot
library(ggsci)    # For additional color scales in ggplot2
```

- Access the datasets
```{r}
diamonds <- diamonds
head(diamonds)  # View the first few rows of the diamonds dataset

# Access the mpg dataset
mpg <- mpg
head(mpg)
```

# 1. Bar Chart
- Used to represent categorical data along the x-axis.

- Basic Bar Plot:
```{r}
# Basic bar plot
ggplot(diamonds, aes(x = cut)) +
  geom_bar() +
  theme_bw()  # Adds a clean theme
```

- Bar Plot with Fill by Categorical Variable:
```{r}
# Bar plot filled by another categorical variable
ggplot(diamonds, aes(x = cut, fill = color)) +
  geom_bar() +
  theme_bw()
```

- Bar Plot with Custom Colors from ggsci:
```{r}
# Using ggsci library for color palette
ggplot(diamonds, aes(x = cut, fill = color)) +
  geom_bar() +
  theme_bw() + scale_fill_manual(values = pal_igv()(7))
```

-Bar Plot with Facets (Split by a Variable):
```{r}
# Split the bar plot into multiple subplots using facet_wrap()
ggplot(data = diamonds, aes(x = cut, fill = clarity)) +
  geom_bar() +
  theme_bw() +
  facet_wrap(~color)
```

# 2. Histogram
- Used to represent the distribution of numerical data on the x-axis, with optional fill by a categorical variable.

- Basic Histogram:
```{r}
# Basic histogram
ggplot(data = diamonds, aes(x = price)) +
  geom_histogram(bins = 30, fill = "grey", color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Price", y = "Count", title = "Graph")
```

- Histogram with Fill by Categorical Variable:
```{r}
# Histogram with fill based on 'cut' category
ggplot(data = diamonds, aes(x = price, fill = cut)) +
  geom_histogram(bins = 30, alpha = 0.5) +
  theme_bw() +
  labs(x = "Price", y = "Count", title = "Graph")
```

# 3. Box Plot
- Used to represent categorical data on the x-axis and numerical data on the y-axis.

- Basic Box Plot:
```{r}
# Basic box plot with drv as categorical and hwy as numeric
ggplot(mpg, aes(drv, hwy)) + theme_bw() + geom_boxplot()
```

- Box Plot with Grouping:
```{r}
# Box plot with grouping by 'drv'
ggplot(data = mpg, aes(x = drv, y = hwy, group = drv)) +
  theme_bw() +
  geom_boxplot()
```

- Box Plot with Fill by Manufacturer:
```{r}
# Box plot with 'manufacturer' as the fill
ggplot(data = mpg, aes(x = drv, y = hwy, fill = manufacturer)) +
  theme_bw() +
  geom_boxplot()
```

- Box Plot with Jitter:
```{r}
# Box plot with jittered points to show distribution
ggplot(data = mpg, aes(drv, hwy)) +
  theme_bw() +
  geom_boxplot() + geom_jitter(width = 0.2, aes(color = trans))
```

# 4. Scatter Plot
- Used to represent two numerical variables for comparison.

- Basic Scatter Plot:
```{r}
# Basic scatter plot with carat vs price
ggplot(diamonds, aes(carat, price)) +
  geom_point() +
  theme_bw()
```

- Scatter Plot with Color Based on Categorical Variable:
```{r}
# Scatter plot colored by 'cut'
ggplot(diamonds, aes(carat, price, color = cut)) +
  geom_point() +
  theme_bw()
```

- Scatter Plot with Another Coloring Method:
```{r}
# Scatter plot with coloring by 'clarity'
ggplot(diamonds, aes(carat, price)) +
  geom_point(aes(color = clarity)) +
  theme_bw()
```

- Scatter Plot with Facets:
```{r}
# Scatter plot divided by 'cut'
ggplot(diamonds, aes(carat, price, color = cut)) +
  geom_point() +
  theme_bw() +
  facet_grid(~cut)
```

- Scatter Plot with Custom Shape:
```{r}
# Scatter plot with custom shape, size, and fill
ggplot(diamonds, aes(carat, price, color = cut)) +
  geom_point(shape = 21, size = 3, stroke = 2, fill = "white") +
  theme_bw() +
  facet_grid(~cut)
```

- Scatter Plot with Linear Trend Line:
```{r}
# Scatter plot with linear regression line
ggplot(diamonds, aes(carat, price, color = cut)) +
  geom_point(shape = 21, size = 3, stroke = 2, fill = "white") +
  theme_bw() +
  stat_smooth(se = F) +  # Add linear trend line
  facet_grid(~cut)
```