---
title: "Manipulating Data with `dplyr`"
author: Mostafa S. Abd El-Maksoud
output: html_notebook
---

- In this notebook, we will learn how to perform data manipulation — and create data processing pipelines — using a powerful package called `dplyr`.

`dplyr` is part of [`tidyverse`](https://www.tidyverse.org/), a collection of R packages for data science. `tidyverse` also includes `ggplot2`, one of the most widely used data visualization packages.

## Preliminaries

- `tidyverse` is not built into R, so we first have to install it:

```{r}
install.packages("tidyverse")
```

- Afterwards, we have to tell R that we want to use `tidyverse`:

```{r}
library("tidyverse")
```

## Goodbye Data Frames! Hello Tibbles

- `tidyverse` introduced a faster and better version of R's built-in data frame; we call this a **tibble**. 

- Since this notebook aims to familiarize ourselves with `dplyr`, we will make a shift from data frames to tibbles. 

- We start by loading our dataset 
- note that tibbles are loaded using `read_delim` while data frames are loaded using `read.delim`:

```{r}
data <- read_delim("data.tsv")
```
We check the columns of our dataset:

```{r}
str(data)
```

We view our dataset (this opens a new tab in RStudio): 

```{r}
View(data)
```

## Selecting Columns & Filtering Rows

- To select columns, we use `select()`. The first argument is the dataset, and the succeeding arguments are the columns to be included.

```{r}
data_subset <- select(data, Accession, Family, Order, Class)
data_subset
```

- To filter rows, we use `filter()`.

```{r}
data_subset <- filter(data_subset, Family != "Unclassified" & Order != "Unclassified" & Class != "Unclassified")
data_subset
```

## Pipe: The "Then" Operator

- The processing that we just performed — selecting columns then filtering rows — is actually a simple pipeline already! But notice how our code can easily become cluttered if we are to add more intermediate steps.

- Fortunately, `dplyr` provides a convenient operator called a pipe: `%>%` (a shortcut to typing this operator is by pressing Ctrl+Shift+M or Cmd+Shift+M for Mac). We can think of `%>%` as equivalent to the English "then."

- To illustrate its usage, we rewrite our pipeline like so: 

```{r}
data_subset_using_pipe <- data  %>% 
  select(Accession, Family, Order, Class) %>% 
  filter(Family != "Unclassified" & Order != "Unclassified" & Class != "Unclassified")
data_subset_using_pipe
```

- Observe how the syntax is mostly the same, with the exception of the first argument of `select()` and `filter()`. Since we already specified `data` at the start of the pipeline, we do not need to pass it anymore as an argument to the data manipulation functions.

## Adding Columns (Mutate)

To add columns, we use `mutate()`.

- We have a column called `Genome Length (bp)` but we want a new column where the genome length is expressed in terms of kbp.

```{r}
# We enclose column names with spaces in backticks ``

data_with_new_column <- data %>%
    mutate(`Genome Length (kbp)` = `Genome Length (bp)` / 1000)
data_with_new_column
```

## Getting a Column (Pull)

- Suppose we want to get all the accessions of the phages in our dataset.
- As we learned earlier, we can use `select()`.

```{r}
data %>% select(Accession)
```

- This works, but this is a one-column tibble. What if we want a vector? We can use `pull()` instead.

```{r}
data %>% pull(Accession)
```
## Split-Apply-Combine Data Analysis

- Let us now try to create more complex pipelines for exploratory data analysis — and, along the way, introduce some functions for aggregating entries and statistics!

- Suppose we want to get the number of phages infecting each host genus. This is a good use case for `group_by()` and `n()` (for counting the number of entries in a group):

```{r}
data %>% 
  select(Accession, Host) %>% 
  filter(Host != "Unspecified") %>%
  group_by(Host) %>% 
  summarize(count = n()) %>%
  arrange(desc(count))           # desc() means arrange in descending order
```

- Suppose we want to get the mean and the median guanine-cytosine (GC) content of the phages when grouped by host genus:

```{r}
# We enclose column names with spaces in backticks ``

data %>% 
  select(Accession, Host, `molGC (%)`) %>% 
  filter(Host != "Unspecified") %>%
  group_by(Host) %>% 
  summarize(mean_gc = mean(`molGC (%)`),
            median_gc = median(`molGC (%)`)) %>%
  arrange(desc(mean_gc))           # desc() means arrange in descending order
```